Titeln på avhandlingens språk

### 2023-05-18 body/abstract.tex
final version

    Next, an exploration of the requirements for realism in the emulation is presented.
    Further, this work discusses the extent to which realism in the emulation can open new avenues for optimization of these systems.
    To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
    The model is combined with a mathematical framework to study the potential for optimization in these applications.

@@

     Den här avhandlingen tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
     Först introduceras ett metodologiskt tillvägagångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
     Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillvägagångssätt eller analytisk modellering.
     Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att bibehålla realismen i nätverks- och datoreffekter.

     Därefter presenteras en utforskning av kraven på realism i emuleringen.
     Vidare diskuterar detta arbete i vilken utsträckning realism i emuleringen kan öppna nya vägar för optimering av dessa system.
     För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobilt förstärkt verklighets-applikationer (engelska \emph{\gls{MAR}}).
     Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i dessa applikationer.

     Resultaten indikerar att den metod som introducerats i detta arbete erbjuder fördelar jämfört med befintliga metoder genom att förbättra effektiviteten, repeterbarheten och replikerbarheten.
     Genom att helt integrera arbetsbelastningskomponenter i den emulerade mjukvarudomänen, minskar denna metodik komplexiteten samtidigt som den fångar komplexa effekter av nätverks- och beräkningsfaktorer som är utmanande att modellera.
     Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
     Dessutom tyder resultat från utforskningen av implikationerna av realism i emuleringen att inkorporering av förbättrad realism i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
     Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

>>>>

    Next, an exploration of the requirements for accuracy in the emulation is presented.
    Further, this work discusses the extent to which accuracy in the emulation can open new avenues for optimization of these systems.
    To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
    The model is combined with a mathematical framework to study the potential for optimization in these applications.

@@

    Denna avhandling tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
    Först introduceras ett metodologiskt tillvägagångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
    Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillvägagångssätt eller analytisk modellering.
    Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att bibehålla realismen i nätverks- och datoreffekter.

    Därefter presenteras en utforskning av kraven på noggrannhet i emuleringen.
    Vidare diskuterar detta arbete i vilken utsträckning noggrannhet i emuleringen kan öppna nya vägar för optimering av dessa system.
    För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobilt förstärkt verklighets-applikationer (engelska \emph{\gls{MAR}}).
    Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i dessa applikationer.

    Resultaten indikerar att den metod som introducerats i detta arbete erbjuder fördelar jämfört med befintliga metoder genom att förbättra effektiviteten, repeterbarheten och replikerbarheten.
    Genom att helt integrera arbetsbelastningskomponenter i den emulerade mjukvarudomänen, minskar denna metodik komplexiteten samtidigt som den fångar komplexa effekter av nätverks- och beräkningsfaktorer som är utmanande att modellera.
    Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
    Dessutom tyder resultat från utforskningen av implikationerna av noggrannhet i emuleringen att inkorporering av förbättrad noggrannhet i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
    Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

### 2023-05-18 body/acks.tex
final version

There are many, many people I owe gratitude to for making this dissertation possible.
I would like to begin by thanking my supervisor, Prof. James Gross.
This dissertation would not have been possible without his guidance and motivation during this long process.
His supervision opened doors that I had never imagined would open for me, and his dry German humor has certainly helped cheer me up whenever I was overwhelmed by some deadline.
Above all, I would like to thank James for his patience and understanding whenever I encountered difficulties, and for the flexibility he has afforded me in completing this Ph.D.

@@

This dissertation would not have been possible without the help and support of my family.
First and foremost, I would like to thank the love of my life and wife~\footnote{\emph{Future-wife} at the time of writing, but \emph{wife} at the time of the defense of this thesis!}, Andresa.
I could not have done this without her companionship and love.
I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.

I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
They have been my rocks, my pillars, since the day I was born, and I am proud to be their son.
A big thank you goes out to my sister, Paola, as well.
Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.
Likewise, I extend my deepest gratitude to everyone in my immediate and extended family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
I would like to make a special mention to my late grandfather, Caupolicán ``Poli'' Muñoz, who always inspired and fanned the flames of academic curiosity in me.
I hope he would have been proud seeing how far I have come.
Another special mention goes to my uncle Manuel, who graciously took me in and gave me a place to stay at the first few months of my Ph.D. studies.

I would also like to thank all of my friends who have made this long, ardous journey more amenable.
I would particularly like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jennifer, Alyson, Kuai-Kuai, Carmen, Moataz and Nora, Eric, and many others.
They made Pittsburgh feel like home, even before I finally moved there permanently in 2022.
I also want to thank Ann-Charlotte in Stockholm, for always being so kind, so considerate, and so much fun to be around.
Further, I am deeply grateful to my friends in Chile who have remained a source of support, even though we only get to see each other once a year when I visit.
This includes, but is not limited to, Juan Pablo, Fernanda, Diego, Camila, George, Patricio, Kevin, Pablo, and many others.
Yet another special mention goes to Diego's family, who have always treated me as one of their own.

Finally, I would be remiss in not mentioning my pets and animal friends who have provided me with love and warmth during these years.
I thank my dogs, Aros and Skatt, who I sadly had to leave behind in Chile with my sister, but who still greet me with enthusiastically wagging tails whenever I visit.
And I thank my cats, Mochi and Tiramisú, who, although have only been in my life for about a year, have sweetened and made much more bearable the long days and nights I have spent writing this dissertation.

\bigskip

\noindent{}Manuel Olguín Muñoz\\
Pittsburgh, May 2023

>>>>

There are many people I owe gratitude to for making this dissertation possible.
I would like to begin by thanking my supervisor, Prof. James Gross.
This dissertation would not have been possible without his guidance and motivation during this long process.
His supervision opened doors that I had never imagined would open for me, and his dry German humor has certainly helped cheer me up whenever I was overwhelmed by some deadline.
Above all, I would like to thank James for his patience and understanding whenever I encountered difficulties, and for the flexibility he has afforded me in completing this Ph.D.

@@

This dissertation would not have been possible without the help and support of my family.
First and foremost, I would like to thank the love of my life and wife, Andresa.
I could not have done this without her companionship and love.
I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.
I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
They have been my rocks, my pillars, since the day I was born, and I am proud to be their son.
A big thank you goes out to my sister, Paola, as well.
Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.
Likewise, I extend my deepest gratitude to everyone in my immediate and extended family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
I would like to make a special mention to my late grandfather, Caupolicán ``Poli'' Muñoz, who always inspired and fanned the flames of academic curiosity in me.
I hope he would have been proud seeing how far I have come.

I would also like to thank all of my friends who have made this long, arduous journey more amenable.
I would specially like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jenn, Alyson, Carmen, Kuai-Kuai and Eric, Moataz and Nora, Russell, and many others.
They made Pittsburgh feel like home, even before I finally moved there permanently in 2022.
Andresa and I are particularly indebted to Jenn and Alyson for helping organize our elopement ceremony during a very stressful period of time, and for always stepping up whenever we need a cat-sitter.
I also want to thank Ann-Charlotte in Stockholm, for always being so kind, so considerate, and so much fun to be around.
Further, I am deeply grateful to my friends in Chile who have remained a source of support, even though we only get to see each other once a year when I visit.
This includes, but is not limited to, Juan Pablo, Fernanda, Diego, Camila, George, Patricio, Kevin, Pablo, and many others.
Yet another special mention goes to Diego's family, who have always treated me as one of their own.

Finally, I would be remiss in not mentioning my pets and animal friends who have provided me with love and warmth during these years.
I thank my dogs, Aros and Skatt, who I sadly had to leave behind in Chile with my sister, but who still greet me with enthusiastically wagging tails whenever I visit.
And I thank my cats, Mochi and Tiramisú, who, although they have only been in my life for about a year, have sweetened and made much more bearable the long days and nights I have spent writing this dissertation.

\medskip
{\begin{flushright}
    Manuel Olguín Muñoz\\
    May 2023
\end{flushright}}

### 2023-05-18 body/part1/chapter1-introduction/introduction.tex
final version

The current model for distributed computing, known as Cloud Computing, allows users to access shared pools of resources, services, and applications over the internet~\cite{gai2012towards}.
Cloud computing offers many advantages, including as virtually unlimited storage and processing power
This achieved by virtue of employing economies of scale in massive installations containing up to thousands of compute servers and networking equipment.
These installations are known as \emph{datacenters}, and often serve massive amounts of users.

@@

Next, we study how improved realism in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
To achieve this, we develop a novel model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
We then combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications.
We find that improved realism in the emulation enables otherwise inviable optimization approaches.

>>>>

The current model for distributed computing, known as Cloud Computing, allows users to access shared pools of resources, services, and applications over the internet~\cite{gai2012towards}.
Cloud computing offers many advantages, including as virtually unlimited storage and processing power.
This is achieved by virtue of employing economies of scale in massive installations containing up to thousands of compute servers and networking equipment.
These installations are known as \emph{datacenters}, and often serve massive amounts of users.

@@

Next, we study how improved accuracy in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
To achieve this, we develop a novel model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
We then combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications.
We find that improved accuracy in the emulation enables otherwise inviable optimization approaches.

### 2023-05-18 body/part1/chapter2-relatedwork/key_contributions.tex
final version

\begin{figure}
    \centering
    \begin{forest}
        for tree={% style of tree
            align=center,
            grow=0,reversed, % tree direction
            parent anchor=east,
            child anchor=west, % edge anchors
            edge={line cap=round},
            outer sep=+1pt, % edge/node connection
            rounded corners,minimum width=15mm,minimum height=8mm, % node shape
            l sep=7mm % level distance
        }
        [{\scshape Contributions}
            [Methodology
                [{Use-case\\studies} [\acs{WCA}] [\acsp{NCS}]]
                [{Tooling} [{EdgeDroid\\(\acs{WCA})}] [{\acs{CLEAVE}\\(\acsp{NCS})}] [Ainur\\(testbed)]]
            ]
            [{Implications\\of Realism}
                [{Characterization of\\human behavior in \acs{WCA}}]
                [{Realistic human\\model for \acs{WCA}}]
                [Exploration of avenues\\for optimization of \acs{WCA}]
            ]
        ]
        \end{forest}
    \caption{Hierarchical overview of the main and secondary contributions of this dissertation.}\label{fig:contribs}
\end{figure}

This thesis presents two key contributions and a number of sub-contributions to literature, as illustrated in \cref{fig:contribs}.
Firstly, we introduce a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
This methodology is validated by example through two representative use-cases, and we provide tooling for its implementation.
Secondly, we study the implications improved client-side emulation realism has on the results produced with our methodology.
These implications are again explored by example.
We design and implement of a detailed model of human user behavior in \gls{WCA}, and then use it to study the potential for optimization of these systems.

@@

The second key contribution of this dissertation relates to the exploration of the question of to which degree a client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing needs to capture certain details of the physical world.
We explore this interrogative through the elaboration of a detailed model of human user behavior in \gls{WCA}, which we then combine with mathematical framework to study the potential for optimization of these applications.

>>>>

\begin{figure}
    \centering
    \begin{forest}
        for tree={% style of tree
            align=center,
            grow=0,reversed, % tree direction
            parent anchor=east,
            child anchor=west, % edge anchors
            edge={line cap=round},
            outer sep=+1pt, % edge/node connection
            rounded corners,minimum width=15mm,minimum height=8mm, % node shape
            l sep=7mm % level distance
        }
        [{\scshape Contributions}
            [Methodology
                [{Use-case\\studies} [\acs{WCA}] [\acsp{NCS}]]
                [{Tooling} [{EdgeDroid\\(\acs{WCA})}] [{\acs{CLEAVE}\\(\acsp{NCS})}] [Ainur\\(testbed)]]
            ]
            [{Implications of\\Accuracy in\\the Emulation}
                [{Characterization of\\human behavior in \acs{WCA}}]
                [{Realistic human\\model for \acs{WCA}}]
                [Exploration of avenues\\for optimization of \acs{WCA}]
            ]
        ]
        \end{forest}
    \caption{Hierarchical overview of the main and secondary contributions of this dissertation.}\label{fig:contribs}
\end{figure}

This thesis presents two key contributions and a number of sub-contributions to literature, as illustrated in \cref{fig:contribs}.
Firstly, we introduce a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
This methodology is validated by example through two representative use-cases, and we provide tooling for its implementation.
Secondly, we study the implications improved client-side emulation accuracy has on the results produced with our methodology.
These implications are again explored by example.
We design and implement of a detailed model of human user behavior in \gls{WCA}, and then use it to study the potential for optimization of these systems.

@@

The second key contribution of this dissertation relates to the exploration of the question of what degree of detail about the physical world must be captured by a client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
We explore this interrogative through the elaboration of a detailed model of human user behavior in \gls{WCA}, which we then combine with mathematical framework to study the potential for optimization of these applications.

### 2023-05-18 body/part1/chapter3-contributions/methodology.tex
final version

In \cref{summary:methodology}, we discuss our methodology for evaluating the performance of latency-sensitive, feedback-loop-based applications enabled by Edge Computing.
This methodology involves emulating only the end-user component of such applications while using the real network and compute components of the system. 
We further discuss the validation of this approach through two representative use-cases for \gls{XR} and \glspl{CPS}.

Next, in \cref{summary:realism}, we explore the extent to which our client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing needs to capture certain details of the physical world.
We broach this question through the elaboration of a detailed model of human behavior in \gls{WCA}, which we then employ in our methodology, in conjunction with a mathematical framework, to study the the potential for optimization of \gls{WCA} applications.

>>>>

In \cref{summary:methodology}, we discuss our methodology for evaluating the performance of latency-sensitive, feedback-loop-based applications enabled by Edge Computing.
This methodology involves emulating only the end-user, client-side component of such applications while using the real network and compute components of the system. 
We further discuss the validation of this approach through two representative use-cases for \gls{XR} and \glspl{CPS}.

Next, in \cref{summary:realism}, we explore the extent of the need for accuracy in our client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
We broach this question through the elaboration of a detailed model of human behavior in \gls{WCA}, which we then employ in our methodology, in conjunction with a mathematical framework, to study the the potential for optimization of \gls{WCA} applications.

### 2023-05-18 body/part1/chapter3-contributions/wca-implications.tex
mark as prelim
final version

\section{Exploring the need for realism in client-side emulation}\label{summary:realism}

In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
In particular, we wish to explore the impact improving the realism of the client-side emulation has on the potential for optimization of the applications under evaluation.
We approach this question by developing an improved, more detailed version of our model of human behavior in \gls{WCA} used in the EdgeDroid tool, which we then incorporate into our methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.

@@

Once we have established a more accurate model of human behavior, we can explore its implications for the optimization of \gls{WCA} through our methodology.
By improving the realism of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
This research highlights the importance of the realism of emulations as  crucial factor in the application of our methodology.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=.6\textwidth]{publications/2021ImpactDelayedResponse/Fig4c}
    \caption{%
        Structure of a \emph{block} in the experimental LEGO task.
        Figure originally published in \cref{paper:olguinmunoz2021impact}.
    }\label{fig:stepblock}
    \todo[inline]{This figure may need reworking for clarity.}
\end{figure}

@@

\subsubsection{Design of a realistic model for human timings in \gls{WCA}}\label{sec:improved_model}

>>>>

\section{Exploring the need for accuracy in client-side emulation}\label{summary:realism}

In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
In particular, we wish to explore the impact improving the accuracy of the client-side emulation has on the potential for optimization of the applications under evaluation.
We approach this question by developing an improved, more detailed version of our model of human behavior in \gls{WCA} used in the EdgeDroid tool, which we then incorporate into our methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.

@@

Once we have established a more accurate model of human behavior, we can explore its implications for the optimization of \gls{WCA} through our methodology.
By improving the accuracy of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
This research highlights the importance of the accuracy of emulations as  crucial factor in the application of our methodology.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=.6\textwidth]{publications/2021ImpactDelayedResponse/Fig4c}
    \caption{%
        Structure of a \emph{block} in the experimental LEGO task.
        Figure originally published in \cref{paper:olguinmunoz2021impact}.
    }\label{fig:stepblock}
    % \todo[inline]{This figure may need reworking for clarity.}
\end{figure}

@@

\subsubsection{Design of a more accurate model for human timings in \gls{WCA}}\label{sec:improved_model}

### 2023-05-18 body/part1/chapter4-conclusions/conclusions.tex
final version

Secondly, we have explored the implications of improved realism on the client-side emulation on the results produced with our methodology.
To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
We have then used this model to study the potential for optimization of these systems.
Our findings indicate that incorporating improved realism in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
These behaviors play a critical role in the overall performance of such applications.

>>>>

Secondly, we have explored the implications of improved accuracy on the client-side emulation on the results produced with our methodology.
To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
We have then used this model to study the potential for optimization of these systems.
Our findings indicate that incorporating improved accuracy in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
These behaviors play a critical role in the overall performance of such applications.
